Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-15 submission/manuscript.Rmd
define 5CKO in main body

We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on male cells[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. Sex is particularly pertinent in the case of KDM5C because it partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022].

>>>>

We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO (5CKO) brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on male cells[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. Sex is particularly pertinent in the case of KDM5C because it partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO, XY 5CKO), female homozygous knockout (XX _Kdm5c_-KO, XX 5CKO), and female heterozygous knockout (XX _Kdm5c_-HET, XX 5CHET) EpiLCs [@samantaActivationXistEvolutionarily2022].
